---
title: "Final Project"
subtitle: "Water Temperature"
author: "Zhanylai Turatkhan kyzy, Julia Kagiliery, Yilun Zhu"
output: pdf_document
geometry: margin=2.54cm
---

**Introduction**


**Motivation/relevance of the study**


**Objectives**


**Dataset Information**


**Methodology / Analysis**


**Description**
1. SARIMA model
2. TBATS model
3. ETS model
4. 
5. SSES model

```{r, echo=FALSE}
library(ggplot2)
library(readxl)
library(cowplot)
library(tidyverse)
library(lubridate)
library(dplyr)
library(knitr)
library(caret)
library(forecast)
library(zoo)
```

```{r, echo=FALSE}
TempData <- read_excel("Temperature2011-2021 (1).xlsx")
TempData <- TempData |>
  mutate(
    `Decimal Years` = date(date_decimal(`Decimal Years`))) |>
  na.omit() 
  
  colnames(TempData) <- c("Date", "Temp", "SD")
```

```{r, echo=FALSE}
head(TempData)
```

```{r, echo=FALSE}
set.seed(123)

index <- createDataPartition(TempData$Temp, p=0.8, list=FALSE)
trainingData <- TempData[index, ]
testData <- TempData[-index, ]
```

```{r, echo=FALSE}
ggplot(trainingData, aes(x=Date, y=Temp)) +
  geom_point() +
  theme_minimal() +
  labs(title = "Temperature over time",
       x = "Date",
       y = "Temperature (°C)")
```
The plot illustrates the temperature variations over time. Here we can see potential seasonal patterns in the vertical alignment of points across the years.

```{r, echo=FALSE}
par(mar=c(3,3,3,0));par(mfrow=c(1,2))
ACF_plot <- Acf(trainingData$Temp, lag=104, main="Autocorrelation of Temperature")
PACF_plot <- Pacf(trainingData$Temp, lag=104, main="Partial autocorr. of Temperature")
```

The clear seasonal pattern in the ACF plot suggests that any time series 
forecasting models should account for seasonal effects, likely including 
seasonal differencing at lag 52 to remove the seasonal component.
The PACF plot suggests considering a few autoregressive terms in the model, 
as indicated by the initial significant lags. The insignificance of the PACF 
after the first few lags suggests that a simple AR model may be sufficient for 
the non-seasonal part of the model, whereas the significant ACF at seasonal 
lags points to the need for a seasonal AR component.
```{r, echo=FALSE}
print(length(TempData$Temp))
```
```{r, echo=FALSE}
# Assuming TempData has columns Date and Temp
TempData$Date <- as.Date(TempData$Date)
zoo_ts <- zoo(TempData$Temp, order.by = TempData$Date)

# Check the structure
print(zoo_ts)
```

```{r, echo=FALSE}
start(zoo_ts)
end(zoo_ts)
```
```{r, echo=FALSE}
split_index <- floor(0.8 * length(zoo_ts))
train_set <- zoo_ts[1:split_index]
test_set <- zoo_ts[(split_index + 1):length(zoo_ts)]
```

```{r, echo=FALSE}
print(head(train_set))
print(head(test_set))
```

```{r, echo=FALSE}
start(train_set)
end(train_set)
```

```{r, echo=FALSE}
start(test_set)
end(test_set)
```


```{r, echo=FALSE}
fit <- auto.arima(temperature_ts, seasonal = TRUE)
```

```{r, echo=FALSE}
print(fit)
```

