---
title: "Final Project"
subtitle: "Water Temperature"
author: "Zhanylai Turatkhan kyzy, Julia Kagiliery, Yilun Zhu"
output: pdf_document
geometry: margin=2.54cm
---

```{r, echo=FALSE}
library(ggplot2)
library(readxl)
library(cowplot)
library(tidyverse)
library(lubridate)
library(dplyr)
library(knitr)
library(caret)
library(forecast)
library(zoo)
```

**Introduction**


**Motivation/relevance of the study**


**Objectives**


**Dataset Information**

```{r, echo=FALSE}
#We important the monthly temperature data
file_path <- "~/WaterTemperature_TSA_FinalProject/Monthly Data_Temperature2011-2021.xlsx"
temperature_data <- read_excel(file_path, sheet = "Sheet1")
```

```{r, echo=FALSE}
#Transforming the data into a time series which starts in Jan of 2011
temperature_ts <- ts(temperature_data$Temperature,
                     start = c(2011, 1), frequency = 12)
```

```{r, echo=FALSE}
#Upon initial examination, the data appears to be strongly seasonal with very little trend.
plot(temperature_ts)
```

```{r}
#The ACF and Pacf tell us that there is ovbious and strong seasonality and some auto-regressive component up to about 6 lags.
P1 <-Acf(temperature_ts)
P2 <- Pacf(temperature_ts)
```

HERE WE SHOULD DECOMPOSE THE DATA.


```{r}
#confirming that timeseries transformation works as expected.
str(temperature_ts)
```


In the lines of code below we split the existing data set into a training data set and a test set. 80% of the data is training and the last 20% is used to validate the model.
```{r}
#Making Training and Test sets
split_point <- round(length(temperature_ts) * 0.8)
train_ts <- window(temperature_ts, end=c(2019, 8))
test_ts <- window(temperature_ts, start=c(2019, 9))
```

```{r}
#Confirms Training and Test sets were correctly made
start(train_ts)
end(train_ts)
start(test_ts)
end(test_ts)
```
**Methodology / Analysis**


**Description**
1. SARIMA model
2. TBATS model
3. ETS model
4. SSES model
5. NN

*SARIMA Modeling*
```{r, echo=FALSE}
#SARIMA autofit
sarima_model <- auto.arima(train_ts)
print(summary(sarima_model))
```

```{r}
checkresiduals(sarima_model)
```

```{r}
h <- length(test_ts)
sarima_forecast <- forecast(sarima_model, h=h)
```

```{r}
autoplot(temperature_ts, series = "Original") +
  autolayer(sarima_forecast$mean, series = "SARIMA forecast") +
  ylab("Water Temperature") +
  ggtitle("SARIMA modeling")
```

```{r}
checkresiduals(sarima_forecast)
```

```{r}
forecast_accuracySarima <- accuracy(sarima_forecast, test_ts)
print(forecast_accuracySarima)
```

*STL + ETS Modeling*

```{r}
# STL + ETS
ets_model <- stlf(train_ts, h=h)
```

```{r}
autoplot(temperature_ts, series = "Original") +
  autolayer(ets_model, series = "STL + ETS", PI=FALSE) +
  ylab("Water Temperature") +
  ggtitle("STL + ETS modeling")
```

```{r}
stlf_accuracy <- accuracy(ets_model, test_ts)
print(stlf_accuracy)
```

```{r}
checkresiduals(ets_model)
```

*TBATS Modeling*

```{r}
model_tbats <- tbats(train_ts)
print(summary(model_tbats))
```

```{r}
checkresiduals(model_tbats)
```


```{r}
TBATS_for <- forecast(model_tbats, h=h)

autoplot(train_ts) +
  autolayer(TBATS_for, series = "TBATS Forecast", PI = FALSE) +
  autolayer(test_ts, series = "Test Data", PI = FALSE) +
  autolayer(fitted(model_tbats), series = "Model", PI = FALSE) +
  ylab("Water Temperature") +
  ggtitle("TBATS Modeling")

```
```{r}
TBATS_accuracy <- accuracy(TBATS_for, test_ts)
print(TBATS_accuracy)
```

